Reaching is Better When You Get What You Want: Realtime Feedback of Intended Reaching Trajectory Despite an Unstable Environment

### 2015-06-03 Paper2frontiers/frontiers.tex
mild updates + fixed figure format

\subsection{Intent Extraction}
The well-known motion control structure of \cite{shadmehr1994adaptive} relates arm trajectory, $q$, to desired arm trajectory, $q_d$, and any external disturbance, $E$ using physical parameters of the arm. To show how this model can be algebraically inverted to instead describe desired arm trajectory as a function of arm trajectory and external disturbance, we write it as a torque balance:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=
\underbrace{\overbrace{\tau_{ff}}^{\text{Feedforward}}+\overbrace{\tau_{fb}}^{\text{Feedback}}}_\text{Controller}
\end{equation}
Where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocities, $\ddot{q}$ is joint angular accelerations, and $G$ contains both Coriolis and centripetal effects. Typical applications solve this torque balance for $\ddot{q}$ and use a numerical differential equation solver to predict arm trajectory in the context of a disturbance of interest, a feedback model, and feedforward torques determined by inverse dynamics. Rather than test hypotheses regarding the learning, production, or composition of this feedforward torque, we instead solve for it:
\begin{equation}
\tau_{ff}=M(q)\ddot{q}+G(q,\dot{q})+E-\tau_{fb}
\end{equation}
Then we note that feedforward torque can have a one-to-one correspondence with desired acceleration, $\ddot{q}_d$:
\begin{equation}
\hat{M}(q_d)\ddot{q}_d+\hat{G}(q_d,\dot{q}_d)+\hat{E}=\tau_{ff}
\end{equation}
Hats ( $\hat{}$ ) denote the nervous system's best estimate of a physical quantity. Combining these expressions, suppressing state dependencies, and solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}^{-1} \{M\ddot{q}+G-\hat{G}+E-\hat{E}-\tau_{fb}\}
\end{equation}
In this form, a differential equation solver can determine $q_d$ as it evolves in time if a few assumptions are made and conditions are met. First, $\hat{E}$ must be modeled or assumed, so we choose $\hat{E}=0$. In the presence of a zero mean white noise force disturbance its mean should be zero, but it is unlikely to be exactly zero and may reflect an average of only the last few exposures \citep{scheidt2001learning}. Next, the matrix $\hat{M}(q_d)$ must be invertible, but we can ensure this through our choice of workspace. Finally, feedback torque requires a model of arm impedance, which is known to be task-dependent \citep{gomi1998task} and may vary over the course of a reach \citep{niu2010temporal}. With no prior knowledge of arm impedance for this task-disturbance combination, we presumed the feedback torque model of Shadmehr and Mussa-Ivaldi anticipating that it is sufficiently accurate or easy to learn (experiment 1). The experiment was repeated with a lower stiffness estimate (experiment 2) to explore any dependence on this assumption.

\subsection{Apparatus}
A planar manipulandum (described in Patton and Mussa-Ivaldi \cite{patton2004robot}) was programmed to minimize any friction or mass. The MATLAB XPC-TARGET package \citep{MATLAB:2008} was used to render this force environment at 1000 Hz and data were collected at 1000 Hz.  Visual feedback of cursor position was performed at 60 Hz using OpenGL. Closed-loop data transmission time (position measurement to completed rendering to recognition of rendering by the position measurement system) was less than 8 milliseconds, ensuring a visual delay less than one 60 Hz frame. Numerical simulation was performed in real-time using the GNU Scientific Library's odeiv2 driver with Runge-Kutta-Fehlberg (4,5) stepping\cite{gough2009gnu}.

\subsection{Human Subjects}
The human data trajectories analyzed here are drawn from nine subjects who gave informed consent in accordance with Northwestern University Institutional Review Board, which specifically approved this study and follows the principles expressed in the Declaration of Helsinki. Seven male and one female right-handed subjects (ages 21 to 30) performed the reaches with their right arm and were not compensated. Subjects' arm segment lengths were directly measured \textit{in situ} while body mass was self-reported.

\subsection{Experimental Design}
We chose 15 centimeters long, center-out reaches of the right arm to visually-presented targets chosen at 60 degree intervals about that central point. Target selection was carried out pseudorandomly such that each outer target was visited 16 times in each of 5 blocks of 96 reaches each. During blocks 2 through 4, subjects experienced filtered white noise forces drawn from a white noise generator at 1000 Hertz with flat power spectral density of 1 Newton, and then passed through a 4th order low-pass Butterworth filter with cutoff $10 \pi$ radians per second. During block 3, cursor position indicated deduced intent. During all other blocks, the cursor position indicated hand position (Figure 2A).

\subsection{Dynamic simulation of arm and intended trajectories}
Anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics} were used to estimate relationships between body mass and limb mass, limb length and limb center of mass, and limb mass and length and moment of inertia. Viscosity parameters, $K_d$, were taken from Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive}. Stiffness parameters were either taken from Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} ($K_{P1}$, experiment 1) or estimated ($K_{P2}$, experiment 2). Expressed in Newton-Meters per Radian:
\begin{equation}
K_{P1}=
\begin{bmatrix}
15 & 6 \\
6 & 16
\end{bmatrix}
\quad
K_{P2}=
\begin{bmatrix}
8 & 2 \\
2 & 5
\end{bmatrix}
\end{equation}
To estimate this reduced stiffness, a pilot subject was asked to intend to remain still on the center target, $q_d$, while co-contracting as little as possible. $K_{P2}$ was then calculated from one minute of white noise forces, $E$, and joint angle traces, $q$ as the least squares solution to the system:
\begin{equation}
K_{P2}\big(q_d(t)-q(t)\big)=M\big(q(t)\big)\ddot{q}(t)+G(q(t),\dot{q}(t))+E(t)+K_D\dot{q}(t)
\end{equation}
where $K_{P2}$ is a 2-by-2 matrix while the state difference and torque are 2-by-60000 matrices.

\subsection{Metrics and Statistical Analysis}
Trajectories and forces were rotated such that movement and force parallel to the line connecting the previous target (the reach origin) and the presented target lie along a \textit{progress} axis while perpendicular movement and force lie along an \textit{error} axis. Reach onset was detected as the moment the cursor's distance from the center of the previous target first exceeded 1 centimeter. Maximum perpendicular error for a trajectory was the largest error magnitude within 250 milliseconds of reach onset. A scalar stiffness, $k$, was calculated for the \textit{error} axis during this same 250 millisecond time span by linear regression:
\begin{equation}
F_e=m\ddot{e}+b\dot{e}+ke+F_O
\end{equation}
Force ($F_e$) and state ($\ddot{e}$, $\dot{e}$, $e$) were known. Mass ($m$), viscosity ($b$), and stiffness offset ($F_0$) terms were calculated, but discarded. While joint stiffness is usually described as a matrix, instantaneous endpoint stiffness in only the error direction is a scalar. This metric isolates stiffness in the error direction and facilitates statistical comparison between treatments and blocks. The paired Wilcoxon signed rank test was used to detect differences in maximum perpendicular error and stiffness between blocks and treatments at the 5\% significance level using the MATLAB statistics toolbox package \cite{MATLAB:2014}.
\end{methods}

@@

\begin{figure}[h]
\centering
\framebox{\parbox{3in}{\includegraphics{Yplots.eps}
}}
\caption{Typical subjects made center-out targeted reaching motions under experimentally varied force and feedback conditions. Subjects used feedback of either hand motions (black lines) or estimated intent (red lines) to complete these reaches. Intent was estimated using either the stiffness model of \cite{shadmehr1994adaptive} (Experiment 1) or a reduced stiffness (Experiment 2) to explore any dependence of reaching stiffness or accuracy on this assumption. The white noise force disturbance was designed to be unpredictable in order to minimize any effect of learning.}
\label{Yplots}
\end{figure}

\begin{figure}[t]
\centering
\framebox{\parbox{3in}{\includegraphics{error.eps}
}}
\caption{Subjects' reaching accuracy depended on the presence of force disturbance and the contents of visual feedback (A). Maximum deviation from straight-line reaching calculated during the first 250 milliseconds after the onset of movement revealed that turbulent force disturbance degraded reaching performance in both hand (B) and intent (C). Comparisons within treatment conditions revealed a significant advantage for the hand when force disturbance is not present and a significant advantage for intent when feedback of intent was used during exposure to turbulent forces. Performance using intent during turbulence was significantly better than performance using the hand (D). Significance among comparisons is denoted by an asterisk. Error did not signficantly depend on the choice between stiffness model of \cite{shadmehr1994adaptive} (experiment 1, blue) or a reduced stiffness (experiment 2, red) to determine intent.}
\label{error}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=10cm]{logo1}% This is a *.jpg file
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures }
\end{figure}

%\begin{figure}
%\begin{center}
%\includegraphics[width=10cm]{logo2}% This is an *.eps file
%\end{center}
%\textbf{\refstepcounter{figure}\label{fig:02} Figure \arabic{figure}.}{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures }
%\end{figure}

\begin{figure}[t]
\centering
\framebox{\parbox{3in}{\includegraphics{stiffness.eps}
}}
\caption{Subjects' arm stiffness depended on the presence of force disturbance and the contents of visual feedback (A). Arm stiffness calculated by linear regression during the first 250 milliseconds after the onset of movement revealed that turbulent force disturbance increased arm stiffness (B). Comparisons between treatment conditions revealed that exposure to turbulent forces caused by arm to stiffen significantly, but feedback of intent could significantly reduce this increase (C). Stiffness did not signficantly depend on the choice between the stiffness model of \cite{shadmehr1994adaptive} (experiment 1, blue) or a reduced stiffness (experiment 2, red) to determine intent.}
\label{stiffness}
\end{figure}

>>>>

\subsection{Intent Extraction}
The well-known motion control structure of \cite{shadmehr1994adaptive} relates arm trajectory, $q$, to desired arm trajectory, $q_d$, and any external disturbance, $E$ using physical parameters of the arm. To show how this model can be algebraically inverted to instead describe desired arm trajectory as a function of arm trajectory and external disturbance, we write it as a torque balance:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=
\underbrace{\overbrace{\tau_{ff}}^{\text{Feedforward}}+\overbrace{\tau_{fb}}^{\text{Feedback}}}_\text{Controller}
\end{equation}
Where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocities, $\ddot{q}$ is joint angular accelerations, and $G$ contains both Coriolis and centripetal effects. Typical applications solve this torque balance for $\ddot{q}$ and use a numerical differential equation solver to predict arm trajectory in the context of a disturbance of interest, a feedback model, and feedforward torques determined by inverse dynamics. Rather than test hypotheses regarding the learning, production, or composition of this feedforward torque, we instead solve for it:
\begin{equation}
\tau_{ff}=M(q)\ddot{q}+G(q,\dot{q})+E-\tau_{fb}
\end{equation}
Then we note that feedforward torque can have a one-to-one correspondence with desired acceleration, $\ddot{q}_d$:
\begin{equation}
\hat{M}(q_d)\ddot{q}_d+\hat{G}(q_d,\dot{q}_d)+\hat{E}=\tau_{ff}
\end{equation}
Hats ( $\hat{}$ ) denote the nervous system's best estimate of a physical quantity. Combining these expressions, suppressing state dependencies, and solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}^{-1} \{M\ddot{q}+G-\hat{G}+E-\hat{E}-\tau_{fb}\}
\end{equation}
In this form, a differential equation solver can determine $q_d$ as it evolves in time if a few assumptions are made and conditions are met. First, $\hat{E}$ must be modeled or assumed, so we choose $\hat{E}=0$. In the presence of a zero mean white noise force disturbance its mean should be zero, but it is unlikely to be exactly zero and may reflect an average of only the last few exposures \citep{scheidt2001learning}. Next, the matrix $\hat{M}(q_d)$ must be invertible, but we can ensure this through our choice of workspace. Finally, feedback torque requires a model of arm impedance, which is known to be task-dependent \citep{gomi1998task} and may vary over the course of a reach \citep{niu2010temporal}. With no prior knowledge of arm impedance for this task-disturbance combination, we presumed the feedback torque model of \cite{shadmehr1994adaptive} anticipating that it is sufficiently accurate or easy to learn (experiment 1). The experiment was repeated with a lower stiffness estimate (experiment 2) to explore any dependence on this assumption.

\subsection{Apparatus}
A planar manipulandum (described in \cite{patton2004robot}) was programmed to minimize any friction or mass. The MATLAB XPC-TARGET package \citep{MATLAB:2008} was used to render this force environment at 1000 Hz and data were collected at 1000 Hz.  Visual feedback of cursor position was performed at 60 Hz using OpenGL. Closed-loop data transmission time (position measurement to completed rendering to recognition of rendering by the position measurement system) was less than 8 milliseconds, ensuring a visual delay less than one 60 Hz frame. Numerical simulation was performed in real-time using the GNU Scientific Library's odeiv2 driver with Runge-Kutta-Fehlberg (4,5) stepping\citep{gough2009gnu}.

\subsection{Human Subjects}
The human data trajectories analyzed here are drawn from sixteen subjects who gave informed consent in accordance with Northwestern University Institutional Review Board, which specifically approved this study and follows the principles expressed in the Declaration of Helsinki. fourteen male and two female right-handed subjects (ages 21 to 30) performed the reaches with their right arm and were not compensated. Subjects' arm segment lengths were directly measured \textit{in situ} while body mass was self-reported.

\subsection{Experimental Design}
We chose 15 centimeters long, center-out reaches of the right arm to visually-presented targets chosen at $60^\circ$ intervals about that central point. Target selection was carried out pseudorandomly such that each outer target was visited 16 times in each of five blocks of 96 reaches each. During blocks 2 through 4, subjects experienced filtered white noise forces drawn from a white noise generator at 1000 Hertz with flat power spectral density of 1 Newton, and then passed through a 4th order low-pass Butterworth filter with cutoff $10 \pi$ radians per second. During block 3, cursor position indicated deduced intent. During all other blocks, the cursor position indicated hand position (Figure 2A).

\subsection{Dynamic simulation of arm and intended trajectories}
Anatomical landmarks and values from \cite{dempster1955space} and \cite{winter2009biomechanics} were used to estimate relationships between body mass and limb mass, limb length and limb center of mass, and limb mass and length and moment of inertia. Viscosity parameters, $K_d$, were taken from Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive}. Stiffness parameters were either taken from \cite{shadmehr1994adaptive} ($K_{P1}$, experiment 1) or estimated ($K_{P2}$, experiment 2). Expressed in Newton-Meters per Radian:
\begin{equation}
K_{P1}=
\begin{bmatrix}
15 & 6 \\
6 & 16
\end{bmatrix}
\quad
K_{P2}=
\begin{bmatrix}
8 & 2 \\
2 & 5
\end{bmatrix}
\end{equation}
To estimate this reduced stiffness, a pilot subject was asked to intend to remain still on the center target, $q_d$, while co-contracting as little as possible. $K_{P2}$ was then calculated from one minute of white noise forces, $E$, and joint angle traces, $q$ as the least squares solution to the system:
\begin{equation}
K_{P2}\big(q_d(t)-q(t)\big)=M\big(q(t)\big)\ddot{q}(t)+G(q(t),\dot{q}(t))+E(t)+K_D\dot{q}(t)
\end{equation}
where $K_{P2}$ is a 2-by-2 matrix while the state difference and torque are 2-by-60000 matrices.

\subsection{Metrics and Statistical Analysis}
Trajectories and forces were rotated such that movement and force parallel to the line connecting the previous target (the reach origin) and the presented target were along a \textit{progress} axis while perpendicular movement and force were along an \textit{error} axis. Reach onset was detected as the moment the cursor's distance from the center of the previous target first exceeded 1 centimeter. Maximum perpendicular error for a trajectory was the largest error magnitude within 250 milliseconds of reach onset. A scalar stiffness, $k$, was calculated for the \textit{error} axis during this same 250 milliseconds time span by linear regression:
\begin{equation}
F_e=m\ddot{e}+b\dot{e}+ke+F_O
\end{equation}
Force ($F_e$) and state ($\ddot{e}$, $\dot{e}$, $e$) were known. Mass ($m$), viscosity ($b$), and stiffness offset ($F_0$) terms were calculated, but discarded. While joint stiffness is usually described as a matrix, instantaneous endpoint stiffness in only the error direction is a scalar. This metric isolates stiffness in the error direction and facilitates statistical comparison between treatments and blocks. The paired Wilcoxon signed rank test was used to detect differences in maximum perpendicular error and stiffness between blocks and treatments at the 5\% significance level using the MATLAB statistics toolbox package \cite{MATLAB:2014}.
\end{methods}

@@

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=10cm]{Yplots.eps}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{ Typical subjects made center-out targeted reaching motions under experimentally varied force and feedback conditions. Subjects used feedback of either hand motions (black lines) or estimated intent (red lines) to complete these reaches. Intent was estimated using either the stiffness model of \cite{shadmehr1994adaptive} (Experiment 1) or a reduced stiffness (Experiment 2) to explore any dependence of reaching stiffness or accuracy on this assumption. The white noise force disturbance was designed to be unpredictable in order to minimize any effect of learning. }
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=10cm]{error.eps}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:02} Figure \arabic{figure}.}{ Subjects' reaching accuracy depended on the presence of force disturbance and the contents of visual feedback (A). Maximum deviation from straight-line reaching calculated during the first 250 milliseconds after the onset of movement revealed that turbulent force disturbance degraded reaching performance in both hand (B) and intent (C). Comparisons within treatment conditions revealed a significant advantage for the hand when force disturbance is not present and a significant advantage for intent when feedback of intent was used during exposure to turbulent forces. Performance using intent during turbulence was significantly better than performance using the hand (D). Significance among comparisons is denoted by an asterisk. Error did not signficantly depend on the choice between stiffness model of \cite{shadmehr1994adaptive} (experiment 1, blue) or a reduced stiffness (experiment 2, red) to determine intent.}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=10cm]{stiffness.eps}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:03} Figure \arabic{figure}.}{ Subjects' arm stiffness depended on the presence of force disturbance and the contents of visual feedback (A). Arm stiffness calculated by linear regression during the first 250 milliseconds after the onset of movement revealed that turbulent force disturbance increased arm stiffness (B). Comparisons between treatment conditions revealed that exposure to turbulent forces caused by arm to stiffen significantly, but feedback of intent could significantly reduce this increase (C). Stiffness did not signficantly depend on the choice between the stiffness model of \cite{shadmehr1994adaptive} (experiment 1, blue) or a reduced stiffness (experiment 2, red) to determine intent.}
\end{figure}
